\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{microtype}

\DeclareUrlCommand{\code}{\urlstyle{tt}}
\newcommand{\bodyterm}[1]{\textbf{#1}}
\emergencystretch=3em
\sloppy
\setlist[itemize]{leftmargin=1.5em}
\setlist[enumerate]{leftmargin=1.5em}
\titleformat{\section}{\Large\bfseries}{\thesection}{0.8em}{}
\titleformat{\subsection}{\large\bfseries}{\thesubsection}{0.8em}{}

\title{Epiphany's Body: A Proprioception Whitepaper}
\author{GameCult / EpiphanyAgent}
\date{June 9, 2026}

\begin{document}
\maketitle

\begin{abstract}
Epiphany is an opinionated Codex-derived agent substrate built around a hard
requirement: the model must model the thing it changes. This whitepaper maps
Epiphany's \bodyterm{Body}: the repository, Rust crates, vendored bridge
machinery, typed state stores, schema catalog, runtime binaries, protocol
surfaces, operator artifacts, and external interface boundaries through which
Epiphany senses and acts. The map is written from the Proprioception faculty:
not as product copy, not as lore, but as an architecture overview whose owner,
input, output, invariant, and cut lines remain inspectable.
\end{abstract}

\section{Definition}

In Epiphany doctrine, the Body is the substrate Epiphany acts through and
senses: repository files, tools, runtime state, prompts, logs, interfaces,
permissions, hosted systems, and editor/runtime bridges. Body is not Mind. Body
is changed and sensed; Mind admits durable state. Proprioception is the faculty
that models the Body: architecture, dataflow, seams, constraints, invariants,
and the difference between the map and the substrate.

The practical definition in this repository is:

\begin{itemize}
  \item the root checkout at \code{E:\Projects\EpiphanyAgent};
  \item Epiphany Rust crates such as \code{epiphany-core}, \code{epiphany-state-model}, model/tool adapters, and OpenAI transport crates;
  \item vendored substrate in \code{vendor/codex}, \code{vendor/cultcache-rs}, \code{vendor/cultnet-rs}, and \code{vendor/cultmesh-rs};
  \item typed state stores under \code{state/};
  \item schemas under \code{schemas/cultnet/};
  \item local operator scripts and generated witness artifacts;
  \item adjacent service/client bodies: Aquarium, Odin/Gjallar, Eve/CultUI, Bifrost, VoidBot, Rider, Unity, Discord, and future GameCult Verse hosts.
\end{itemize}

\section{Executive Map}

Epiphany's current Body is a hybrid machine in transition. Its center of gravity
has moved into typed Rust/CultCache organs, but compatibility arteries still
run through vendored Codex. The live target is not a cleaner pile of Codex
routes. The target is an Epiphany-native CultCache/CultMesh/CultNet body with a
relatively vanilla Codex-derived reliquary retained for OpenAI subscription
auth and model transport.

\begin{longtable}{>{\raggedright\arraybackslash}p{0.23\linewidth}>{\raggedright\arraybackslash}p{0.32\linewidth}>{\raggedright\arraybackslash}p{0.35\linewidth}}
\textbf{Body Region} & \textbf{Current Owner} & \textbf{Invariant} \\
\hline
Durable thread state & \code{epiphany-state-model}, \code{epiphany-core} & State shape is typed; prompt rendering is a projection, not authority. \\
Domain policy & \code{epiphany-core} & Verdicts, launch policy, acceptance rules, runtime lifecycle, and mutation law live here, not in bridge DTOs. \\
Vendored Codex & \code{vendor/codex} & Retain host/app-server/auth/model compatibility where useful; do not let Codex own Epiphany prompts, scheduler, or state. \\
Runtime spine & \code{epiphany-core/src/runtime_spine.rs} and \code{state/runtime-spine.cc} & Runtime identity, jobs, events, receipts, tool/model documents are typed CultCache records. \\
Organ memory & \code{state/agents.cc}; agent-memory CLI & Resident work organs use lean work-organ state; Persona uses portable Persona state. \\
Heartbeat physiology & \code{state/agent-heartbeats.cc} via \code{epiphany-heartbeat-store} & Scheduling, cooldown, rumination, routine, cognition lanes, and active-turn freeze are scheduler physiology, not project truth. \\
Local Verse & CultMesh over \code{state/local-verse.cc} & Local Verse context is compact prompt/operator context; private state does not become public by flipping a flag. \\
Schema catalog & \code{schemas/cultnet/index.json} & JSON schema publishes contracts; typed Rust documents remain runtime authority. \\
Operator artifacts & \code{.epiphany-*}, \code{tools/epiphany_local_run.ps1} & Artifacts witness and lower state; they do not become state owners. \\
\end{longtable}

\section{Repository Body}

The checkout has nine primary Epiphany Rust crates:

\begin{itemize}
  \item \code{epiphany-core}: the main domain organ. It owns state-update
  validation, launch documents, runtime spine, heartbeat, memory graph, Persona
  projection, Substrate Gate, Mind, Eyes, Hands, Soul, Continuity, CultMesh
  integration, surface derivation, smokes, and local operator binaries.
  \item \code{epiphany-state-model}: the durable \code{EpiphanyThreadState}
  type plus \code{render_epiphany_state}. This is the typed Mind surface that
  prompt fragments and views read.
  \item \code{epiphany-model-adapter}: provider-neutral model request, stream
  event, receipt, and status documents.
  \item \code{epiphany-openai-adapter}: OpenAI-shaped request/event/receipt and
  status documents.
  \item \code{epiphany-openai-auth-spine}: Codex-derived auth boundary over
  \code{codex-login}. It preserves honest Codex-compatible credential and token
  behavior rather than impersonating it from scratch.
  \item \code{epiphany-openai-codex-spine}: model transport over vendored Codex
  client machinery.
  \item \code{epiphany-openai-runtime}: executable model runtime, including
  \code{epiphany-model-runtime}.
  \item \code{epiphany-tool-adapter}: typed tool capability, invocation intent,
  and invocation receipt documents.
  \item \code{epiphany-tool-codex-mcp-spine}: quarantined Codex MCP adapter. It
  consumes typed tool invocation intents and emits typed receipts; raw MCP JSON
  remains adapter-edge cargo.
\end{itemize}

The vendored region contains:

\begin{itemize}
  \item \code{vendor/codex}: the Codex-derived host and compatibility substrate.
  It is tracked directly, not as a submodule.
  \item \code{vendor/cultcache-rs}, \code{vendor/cultnet-rs}, and
  \code{vendor/cultmesh-rs}: repo-contained Rust Cult substrate.
\end{itemize}

\section{State Body}

The \code{state/} directory is the local persistent body. The important stores
are:

Canonical CultCache store files use the \code{.cc} extension. The local state
body has been migrated to that naming directly; serialization details are
implementation, not file authority.

\begin{longtable}{>{\raggedright\arraybackslash}p{0.30\linewidth}>{\raggedright\arraybackslash}p{0.60\linewidth}}
\textbf{Store} & \textbf{Role} \\
\hline
\code{state/map.yaml} & Canonical slow project map: objective, constraints, invariants, dataflow, active subgoals, current status, guardrails, and rejected paths. \\
\code{state/ledgers.cc} & CultCache-shaped evidence and branch ledger. Future behavior-changing lessons go here, not in an endless activity feed. \\
\code{state/agents.cc} & Organ self-memory and Persona projection source. Resident work organs use \code{work_organ}; public/person-shaped agents use \code{persona}. \\
\code{state/agent-heartbeats.cc} & Heartbeat initiative, cooldown, pending turns, cognition lanes, routine physiology, and scheduler pressure. \\
\code{state/runtime-spine.cc} & Runtime identity, sessions, jobs, job results, events, worker launch requests, role/reorient findings, receipts, model/tool documents, and schema catalog surfaces. \\
\code{state/thread-state.cc} & Native thread-state store. \\
\code{state/local-verse.cc} & Local CultMesh Verse context for compact operator and prompt-context packets. \\
\code{state/memory-graph.cc} & Local memory graph store. \\
\code{state/persona-discord.toml} & Persona Discord boundary configuration. \\
\code{state/void-memory.toml} & Void memory bridge configuration. \\
\code{state/scratch.md} and \code{state/scratch-compaction-*.md} & Scratch surfaces. Scratch is a workbench, not canonical memory. \\
\end{longtable}

Lock files sit beside several stores. Some local generated state is currently
untracked, including \code{state/local-verse.cc},
\code{state/memory-graph.cc}, and \code{.epiphany-character-loop/}.

\section{CultNet Schema Body}

The schema catalog in \code{schemas/cultnet/index.json} is the public nerve
chart. Current local grouping from the catalog is approximately:

\begin{itemize}
  \item 10 agent-state schemas;
  \item 5 heartbeat schemas;
  \item 21 intent schemas;
  \item 8 model/OpenAI schemas;
  \item 3 tool schemas;
  \item 6 runtime schemas;
  \item 17 surface schemas;
  \item 4 receipt/ledger schemas;
  \item 7 other schemas.
\end{itemize}

The catalog includes runtime identity/session/job/event/result documents, model
requests and receipts, tool invocation intents and receipts, Persona and
work-organ state, heartbeat state, operator CultMesh documents, scene/freshness
/context/graph/pressure/reorient/roles/coordinator surfaces, Unity and Rider
bridge surfaces, Void memory, and repo-initialization documents.

The invariant is simple: schema JSON describes the contract. Load-bearing
runtime state should be typed Rust documents persisted through CultCache or
projected through CultMesh/CultNet. JSON may remain at hostile edges, schema
publication, sealed forensic artifacts, or named quarantine experiments.

\section{Runtime Spine}

\code{epiphany-core/src/runtime_spine.rs} is one of the main spinal cords of
the Body. It imports and registers:

\begin{itemize}
  \item runtime identity, session, job, job-result, and event records;
  \item worker launch requests and role/reorient worker result documents;
  \item model adapter, OpenAI adapter, and tool adapter documents;
  \item surface document types for scene, freshness, context, graph query,
  pressure, reorient, CRRC, jobs, roles, role result, reorient result,
  planning, coordinator, Persona, Void memory, repo initialization, repo birth
  runner, Rider bridge, and Unity bridge;
  \item organ receipts for Mind, Eyes, Hands, Soul, Continuity, and Substrate
  Gate;
  \item CultMesh operator status, snapshots, run intents, run receipts, Verse
  policy, global room policy, and organ contract documents.
\end{itemize}

Runtime state is machinery telemetry. It is not project truth. Durable thread
state may link to runtime records through runtime links, but it must not mirror
or duplicate their lifecycle fields as a competing owner.

\section{Faculties And Protocol Machinery}

The embodied sub-agents are:

\begin{itemize}
  \item \bodyterm{Self}: coordination, routing, authority boundaries.
  \item \bodyterm{Persona}: public and social expression.
  \item \bodyterm{Imagination}: futures, plans, objective drafts, projection.
  \item \bodyterm{Eyes}: research, source inspection, evidence ingress.
  \item \bodyterm{Proprioception}: internal model of Body shape, dataflow,
  architecture, checkpoints, and seams.
  \item \bodyterm{Hands}: commands, edits, commits, rollbacks, external action.
  \item \bodyterm{Soul}: verification, invariants, ethics, falsification.
\end{itemize}

The machinery surfaces are:

\begin{itemize}
  \item \bodyterm{Body}: substrate. It is sensed and changed; it does not decide.
  \item \bodyterm{Mind}: persistent state and state admission.
  \item \bodyterm{Continuity}: compaction, recovery, sleep, stale-turn repair,
  and reorientation receipts.
  \item \bodyterm{Substrate Gate}: scoped repository/substrate access.
\end{itemize}

Every embodied sub-agent depends on the other embodied sub-agents, but
dependency does not collapse ownership. A Persona turn, for example, should
move through Substrate-Gate-scoped facts, Mind state, social stimulus, organ
dependencies, Imagination as Projector, Persona as natural-language actor, and
Mind as Interpreter. None of those steps grants Persona hidden state authority.

\section{Codex Reliquary}

Epiphany publishes no app-server JSON-RPC state surface. Vendored Codex owns
only Codex concerns and the retained OpenAI-compatible auth/model transport
machinery. Epiphany state, prompts, scheduling, interfaces, and durable
invariants remain native typed organs.

The retained Codex body currently matters for:

\begin{itemize}
  \item Codex-compatible auth identity and token refresh through
  \code{codex-login};
  \item OpenAI/Codex model transport through vendored client machinery;
  \item useful app-server and streaming compatibility affordances while
  Epiphany surfaces are still evacuating to native CultNet/CultMesh contracts.
\end{itemize}

The starvation doctrine is the deletion line: Codex may be a bridge and a
reliquary, but not Epiphany's prompt authority, state owner, scheduler, runtime
truth, operator API owner, or plugin marketplace brain.

\section{Operator And Interface Body}

Current operator access is partly native and partly witness-artifact based:

\begin{itemize}
  \item \code{tools/epiphany_local_run.ps1} is the one-command local operator
  wrapper.
  \item Native binaries such as \code{epiphany-mvp-status},
  \code{epiphany-mvp-coordinator}, \code{epiphany-verse-query},
  \code{epiphany-operator-run}, \code{epiphany-operator-snapshot},
  \code{epiphany-eve-surface}, \code{epiphany-runtime-spine},
  \code{epiphany-heartbeat-store}, and \code{epiphany-agent-memory-store}
  inspect or update typed state.
  \item Local artifact directories such as \code{.epiphany-run/},
  \code{.epiphany-dogfood/}, \code{.epiphany-smoke/},
  \code{.epiphany-rumination/}, \code{.epiphany-gui/}, and
  \code{.epiphany-character-loop/} witness work and lower typed state for
  operator inspection.
  \item Aquarium is the richer adjacent operator client. It should render and
  steer typed state, not become a state owner.
  \item Eve/CultUI is the intended GUI/TUI projection mechanism for typed
  CultMesh surfaces.
  \item Odin/Gjallar/Yggdrasil are the intended GameCult Verse discovery and
  hosting organs for daemon affordances and important trusted Verses.
\end{itemize}

The service contract invariant is that status surfaces, dashboards, shell
wrappers, bridge adapters, app-server compatibility layers, and renderers must
read typed projections or emit typed command intents. They must not directly
mutate Epiphany truth.

\section{Primary Dataflows}

\subsection{State Rehydration And Prompt Projection}

Durable thread state lives as typed \code{EpiphanyThreadState}. On re-entry,
state is reconstructed from rollout/native state, then rendered into bounded
prompt-facing context through \code{render_epiphany_state}. Prompt injection is
read-only. If a model learns something from that prompt, admitting it back into
durable state requires an explicit update, proposal, promotion, role acceptance,
or other reviewed write path.

\subsection{View And Surface Projection}

Scene, jobs, freshness, context, graph query, planning, pressure, reorientation,
CRRC, roles, role result, reorient result, coordinator, Persona, bridge, and
environment surfaces are projections. Their job is to make typed state visible
and actionable. A projection does not own its inputs. A dashboard is a window,
not the vault.

\subsection{Worker Launch And Runtime Results}

Launch requests are typed documents. Runtime spine opens jobs, links them to
durable bindings, records model/tool activity, and stores results/receipts.
Role and reorientation findings remain reviewable until Mind and relevant gates
accept them. Implementation now routes through Hands-shaped branch-turn
constraints and should yield one branch/commit evidence slice before
Proprioception refresh.

\subsection{Tool Calls}

Tool use travels through typed capability, invocation intent, and invocation
receipt documents. The Codex MCP adapter is an external compatibility executor.
MCP's raw JSON belongs at that boundary; Epiphany's internal tool state is the
typed intent/receipt pair.

\subsection{Local Verse And Daemon Awareness}

Local Verse context is queried through CultMesh-backed state and can import
Gjallar daemon affordance summaries. The desired daemon path is Odin/Gjallar for
discovery, Eve/CultUI for inspection surfaces, CultMesh/CultNet for state and
commands, and provider-owned receipts for actions. MCP is not the native
GameCult daemon-control path.

\section{Invariants}

\begin{enumerate}
  \item CultCache documents are the data model.
  \item CultNet is the wire for Epiphany-controlled subsystems.
  \item CultMesh is the preferred ergonomic local Verse and operator-store
  layer over raw CultCache/CultNet.
  \item JSON schema may publish the contract; generic JSON must not become
  load-bearing internal state when both sides are Epiphany-owned.
  \item Codex is retained for honest compatibility, especially OpenAI auth and
  model transport. It must not become Epiphany's state/process/prompt owner.
  \item Runtime state is machinery telemetry, not project truth.
  \item Operator artifacts are witnesses, not owners.
  \item Persona is public/person-shaped state; resident work organs use lean
  work-organ state.
  \item Heartbeat is scheduler physiology, not project truth.
  \item Substrate Gate grants access; Eyes inspects evidence; Hands acts; Soul
  verifies; Mind admits durable state; Self routes; Proprioception maps Body.
\end{enumerate}

\section{Current Cut Line}

The Body is coherent enough to inspect and extend. The next cuts are concrete:

\begin{itemize}
  \item Keep \code{thread/epiphany/*} JSON-RPC as compatibility projection
  only. New operator authority belongs in typed CultNet/CultMesh contracts.
  \item \code{codex_message_processor.rs} still participates in routing and
  response projection; move Epiphany-owned routing out through the bridge and
  native runtime surfaces.
  \item Some \code{serde_json::Value} and JSON payloads survive at ingress,
  model output, schema, sealed artifacts, and quarantine seams. Preserve them
  only at those boundaries; typed documents own Epiphany-internal state.
  \item Publish Eve/CultUI compositions from provider-owned CultMesh state;
  renderers lower the composition graph without becoming state owners.
  \item Treat local Verse and memory graph stores as generated runtime state.
  They are inspected through tools and receipts, not tracked as repository
  canon.
  \item The persisted next action after this pass remains the non-ephemeral
  Hands branch-turn smoke and Proprioception refresh proof.
\end{itemize}

\section{Conclusion}

Epiphany's Body is no longer a prompt style pasted onto Codex. It is a growing
typed substrate: Rust domain organs, CultCache stores, CultMesh local Verse
documents, CultNet schema contracts, runtime-spine receipts, heartbeat
physiology, memory graph state, model/tool adapter documents, and local
operator surfaces. The current cut line is split authority: Codex host routes,
JSON-RPC compatibility, JSON cargo, and operator artifacts must remain
compatibility or witness surfaces rather than organs.

The cut line is therefore architectural, not cosmetic. Keep Codex as a
narrow auth/model/compatibility reliquary. Move Epiphany-owned state, scheduler
policy, prompt authority, worker lifecycle, operator surfaces, daemon discovery,
and action receipts through typed CultCache, CultMesh, CultNet, and Eve/CultUI.
That is the Body becoming legible enough to be trusted and sharp enough to keep
building.

\end{document}
